---
title: "State of Women's Soccer 2019"
author: "Bernadette Stadler"
output: pdf_document
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr)
library(ggplot2)
library(plotly)
library(readxl)
library(janitor)
library(reshape2)
library(scales)
library(gt)
library(sf)
library(fs)
library(cowplot)
library(tidyverse)
```

```{r read in data, include=FALSE}
  rev_exp <- read_excel("raw-data/Revenue and expense data.xlsx") %>%
    clean_names()

  # read in U.S. soccer revenue and expenses dataset

  wws <- read_csv("raw-data/tabula-GSSS 2017.csv", col_types = cols(
    league = col_character(),
    country = col_character(),
    Season = col_character(),
    Sport = col_character(),
    Teams = col_double(),
    Players = col_double()
  )) %>% clean_names()

  # read in womens salaries data

  mws <- suppressWarnings(read_csv("raw-data/tabula-GSSS 2017_2.csv", col_types = cols(
    RANK = col_character(),
    TEAM = col_character(),
    LEAGUE = col_character()
  ))) %>%
    drop_na() %>%
    filter(!LEAGUE == "LEAGUE") %>%
    clean_names()

  # read in mens salaries data for 2017.Use suppress warnings because I got a warning
  # saying that a missing column name was filled in with X4 that I couldn't get to
  # to go away. Filter to get rid of multiple table headers from subsequent pages of
  # the document I scraped it from.

  mws_18 <- suppressWarnings(read_csv("raw-data/tabula-GSSS 2018.csv", col_types = cols(
    COUNTRY = col_character(),
    CONTINENT = col_character(),
    X4 = col_character()
  ))) %>% clean_names()

  # read in mens salaries data for 2018.Use suppress warnings for same reason as above

  bonuses <- read_excel("raw-data/World cup bonuses.xlsx") %>% clean_names()

  # read in bonuses data (manually compiled from Guardian article)

  model_data <- read_csv("raw-data/womens-football-2019-clean.csv",
    col_types = cols(
      X1 = col_character(),
      Population = col_number(),
      `Global Gender Gap Ranking 2018` = col_character(),
      `FIFA WF Ranking 2019` = col_character(),
      `Female players playing organized football` = col_character(),
      `licesned coaches` = col_character(),
      `licensed referees` = col_character(),
      `Women's soccer strategy?` = col_character(),
      `Mixed-gender football?` = col_character()
    )
  ) %>% clean_names()
  
  # read in fifa data 

  gdp_data <- read_csv("raw-data/GDP_data.csv",
    skip = 4,
    col_types = cols(
      .default = col_double(),
      `Country Name` = col_character(),
      `Country Code` = col_character(),
      `Indicator Name` = col_character(),
      `Indicator Code` = col_character(),
      `2019` = col_logical(),
      X65 = col_logical()
    )
  ) %>% clean_names()
  
  # read in world bank data 

  gender_index <- read_csv("raw-data/data.csv",
    col_types = cols(
      `Country ISO3` = col_character(),
      `Country Name` = col_character(),
      `Indicator Id` = col_double(),
      Indicator = col_character(),
      `Subindicator Type` = col_character(),
      `2006` = col_double(),
      `2007` = col_double(),
      `2008` = col_double(),
      `2009` = col_double(),
      `2010` = col_double(),
      `2011` = col_double(),
      `2012` = col_double(),
      `2013` = col_double(),
      `2014` = col_double(),
      `2015` = col_double(),
      `2016` = col_double(),
      `2018` = col_double()
    )
  ) %>% clean_names()
  
  # read in gender index data 
  
  reg_season_salary <- read_excel("raw-data/reg season salaries.xlsx") %>% clean_names()
```

In March 2019, the 28 members of the United States Women's National Soccer Team (USWNT) filed a class-action lawsuit against their employer, U.S. Soccer. The lawsuit accuses U.S. soccer of violating the Equal Pay Act and Title VII of the Civil Rights Act of 1964 by consistently paying USWNT athletes less than their male counterparts (players on the U.S. Men's National Team or USMNT). The lawsuit, and the subsequent dominance of the USWNT in the 2019 Women's World Cup, have drawn significant attention to the issue of equal pay in professional soccer. However, the USWNT and USMNT have different pay structures that makes it difficult to compare their earnings. This project seeks to explain the USWNT and USMNT compensation structures and earning potentials in a way that allows for useful comparisons. It will also go beyond the USWNT to examine how professional female soccer players are compensated throughout the world. Please feel free to contact me at bstadler@g.harvard.educ with any questions about this project. 

# Regular Season Pay

The USWNT and USMNT have different payment structures, which makes it different to compare their salaries. USWNT players make a base salary of $100,000 and earn bonuses of on average $4,950 for games they win. The actual amount they earn may depend on the FIFA rating of their opponent. The USMNT does not have a base salary, but players make a bonus of $5,000 for losing a game and up to $17,625 for a win, depending on the opponent's ranking. On average, USMNT players would make around $13,166 per game that they win. The USMNT and USWNT also play differing number of games per year, further complicating efforts to directly compare their salaries. The graph below shows how much the USMNT and USWNT would make if each team played 20 games, depending on the number of games won.

```{r prep graph data, include = FALSE }

graph_data1 <- reg_season_salary %>% 
      pivot_longer(cols = women:men, names_to = "team", values_to = "salary") %>% 
      
      # pivot data so that women or men appear as values in column "team" so that I 
      # can facet wrap by team later 
      
      mutate(team = if_else(team == "women", "Women", "Men")) %>%
      
      # change "men" and "women" to "Men" and "Women" so that they will display 
      # nicely in the legend 
  
      filter(games_won %in% c(0,10,20)) %>%
    
      # filter so that I can show three different outcomes in my static graphic. In my app
      # this is interactive
      
      mutate(games_won = case_when(games_won == 0 ~ "No Games Won", 
                                   games_won == 10 ~ "Ten Games Won", 
                                   TRUE ~ "All Games Won"))
  
      # Changes numbers to something that will make sense as a label in my graph 

graph_data1$games_won <- factor(
  graph_data1$games_won, levels = c("No Games Won", "Ten Games Won", "All Games Won")
  )

      # I manually reorded the chart.
```

```{r reg season graph, echo = FALSE, fig.align = "center", out.width = "4.5in", out.height = "4.5in"}

graph_data1 %>%
      ggplot() +
      geom_col(aes(x = team, y = salary, fill = team), position = "dodge") +
  
      # make a ggplot! Set position to dodge so the columns appear next to
      # eachother 
      
      facet_wrap(~games_won) +
      
      # facet wrap by games won to make three separate graphs 
      
      scale_fill_manual(values = c("Women" = "red", "Men" = "purple4")) +
      
      # manually color the bars, using same color scheme as rest of men/women 
      # graphs 
  
      scale_y_continuous(labels = number_format(scale = 0.001)) +
  
      # put numbers in regular notation (not scientific)
      
      coord_flip() +
      
      # flip coordinates because a two-column graph looks better going 
      # horizontally
      
      theme_minimal() +
      
      # theme of my app! 
      
      theme(axis.text.y = element_blank()) + 
      
      # remove y axis text because fill color already identifies each bar
      
      labs(title = "Money earned per games won", 
           subtitle = "Assuming each team plays 20 games", 
           y = "Money Earned (in thousands of dollars)", 
           x = "", 
           caption = "Data from USWNT lawsuit", 
           fill = "Team") +
  
      # add labels 
  
      theme(plot.title = element_text(hjust= 0.5)) +
      theme(plot.subtitle = element_text(hjust = 0.5))

      # center title and subtitle 


```


# World Cup Pay 

USMNT and USWNT players also make different amounts in bonuses for each stage of the world cup they reach, from being named to the world cup team to winning the world cup. The different bonuses made at each stage add up to quite a difference. This chart shows the cumulative amount that a USWNT or USMNT player would walk away with if their team exited the tournament at each respective stage.

```{r read in and clean data, include = FALSE}

bonuses <- read_excel("raw-data/World cup bonuses.xlsx") %>% clean_names()

# read in data, clean names

bonuses2 <- bonuses %>%
  select(stage, women, men, cumulative_women, cumulative_men) %>%

  # select columns of interest

  pivot_longer(
    cols = c("women", "men"), names_to = "team", values_to = c("Value")
  ) %>%

  # pivot longer so that men and women are in the column team so that I will
  # be able to facet wrap by team later to make two side-by-side graphc

  pivot_longer(cols = c("cumulative_women", "cumulative_men"), names_to = "team2", values_to = "cumulative") %>%

  # pivot longer again to get all of the cumulative data in one column.
  # unfortunatey this second pivot gave me some wierd rows that had both women
  # and men_cumulative or vice versa. So, see next step ...

  mutate(keep = case_when(
    team == "men" & team2 == "cumulative_women" ~ FALSE,
    team == "women" & team2 == "cumulative_men" ~ FALSE,
    TRUE ~ TRUE
  )) %>%

  # make a new column that identifies which rows mix teams with FALSE.

  filter(keep == "TRUE") %>%

  # filter for the rows that did not mixt teams, which I labeled as TRUE
  # in the last step

  select(stage, team, Value, cumulative) %>%

  # select variables of interest

  mutate(team = if_else(team == "women", "USWNT", "USMNT")) %>%

  # change "men" and "women" to "USWNT" and "USMNT" to make the graph look more
  # professional

  mutate(number = case_when(
    stage == "Start" ~ 1,
    stage == "Qualifying" ~ 2,
    stage == "Winning qualification games" ~ 3,
    stage == "Named to world cup team" ~ 4,
    stage == "Advance to knockout" ~ 5,
    stage == "Winning world cup" ~ 6,
    TRUE ~ 7
  ))

# give each stage a number. This allows me to make a line chart below (I was
# having some issues when I was using the discrete charcter values). 

```

```{r bonsuses graph, echo=FALSE, fig.align = "center", out.width = "5in", out.height = "4in"}

ggplot(bonuses2, aes(x = number, y = cumulative)) +
  
  # make the ggplot and save it as world_cup_bonuses
  
  geom_point() +
  
  # add the geom_point layer
  
  geom_line() +
  
  # agg the geom_line layer 
  
  facet_wrap(~team) +
  
  # facet_wrap by team 
  
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7), labels = c("Start", "Qualifying", "Winning qualification games", "Named to world cup team", "Advance to knockout", "Winning world cup", "Victory tour")) +
  
  # manually add breaks at every number (previously it didn't display all of them),   # label each number with its associated stage 
  
  scale_y_continuous(labels = number_format(scale = 1)) +
  
  # change y axis labels so they disply as regular numbers, not scientific notation
  
  labs(x = "Stage",
       y = "Cumulative", 
       title = "Cumulative Bonuses per World Cup Stage", 
       caption = "Data from The Guardian") +
  
  # add labels to x and y axis, title, caption
  
  theme_minimal() +
  
  # theme minimal because that is the style of my app
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

  # tilt tick labels for readability 

```

# Market realities? 

In response to the USWNT lawsuit, U.S. Soccer has claimed that it is justified in paying female athletes less because of the market reality that women's soccer generates less revenue. This graph shows USWNT and USMNT revenue, expenses, and net income for fiscal year 2014 through the projected values for 2019. While the USWNT had generated less income for U.S. soccer in some years, the reverse is true in others.

```{r revenue and expenses data, include=FALSE}
rev <- rev_exp %>%

      # first, I'm just going to work with revenue data

      select(fiscal_year, womens_revenue, mens_revenue) %>%

      # select variables of interest

      pivot_longer(
        cols = ends_with("revenue"),
        names_to = "Team",
        values_to = "Revenue"
      ) %>%

      # pivot the data so that the USMNT/USWNT data ends up in one column
      # this will allow me to facet wrap my graph by men and women later

      mutate(fiscal_year = if_else(
        fiscal_year == "2019 (projected)", 2019, suppressWarnings(as.double((fiscal_year)))
      )) %>%

      # mutate to remove (projected) so that I can make fiscal_year a
      # double column. Use suppressWarnings to get rid of the NA introduced by coercion
      # warning (I couldn't figure out how to do it any other way)

      mutate(Team = if_else(Team == "womens_revenue", "USWNT", "USMNT"))

    # mutate to change womens_revenue and mens_revenue to USWNT and USMNT,
    # which is how I want it to show up in the graph

    exp <- rev_exp %>%

      # now work with expense data

      select(fiscal_year, womens_expenses, mens_expenses) %>%

      # select variables of interest

      pivot_longer(
        cols = ends_with("expenses"),
        names_to = "Team",
        values_to = "Expenses"
      ) %>%

      # pivot the data so that the USMNT/USWNT data ends up in one column
      # this will allow me to facet wrap my graph by men and women later

      mutate(fiscal_year = if_else(
        fiscal_year == "2019 (projected)", 2019, suppressWarnings(as.double(fiscal_year))
      )) %>%

      # make 2019 (projected) into a numeric value, as well as whole column. Use
      # suppress warning for same purposes as above.

      mutate(Team = if_else(Team == "womens_expenses", "USWNT", "USMNT"))

    # mutate mens_expenses and womens_expenses to USWNT and USMNT

    rev_exp_formatted <- exp %>%
      left_join(rev, by = c("fiscal_year", "Team")) %>%

      # rejoin the data

      mutate(Expenses = if_else(is.na(Expenses), 0, Expenses)) %>%

      # change NAs to 0 revenue so that I can collapse the rows. The NAs were
      # introduced because I have separate rows for revenue and expense

      mutate(Revenue = if_else(is.na(Revenue), 0, Revenue)) %>%

      # change NAs to 0 in revenue so that I can add and subtract data below

      mutate(Net = Revenue - Expenses) %>%

      # make a net column by subtracting expenses from revenue

      pivot_longer(
        cols = Expenses:Net,
        names_to = "Type",
        values_to = "Amount"
      )

    # pivot longer so that I will be able to filter by Revenue, Expenses and Net.
    # I'm not sure if this is by design, but this is one of the few ways I was able
    # to get reactive variables to work in shiny

    rev_exp_formatted$Type <- factor(rev_exp_formatted$Type, levels = c("Revenue", "Expenses", "Net"))

    # assign levels to Revenue, Expenses and Net so that they display with Net last
    # instead of in alphabetical order
```

```{r revenue and expenses graph, echo=FALSE, fig.align = "center", out.height= "4in", out.width= "5in"}

    rev_exp_formatted %>%
      # make a ggplot

      ggplot(aes(x = fiscal_year, y = Amount, fill = Type)) +

      # fiscal_year goes on x axis, amount on y axis, color by type

      geom_col(position = "dodge") +

      # set columns to dodge so that they display side by side instead of stacked

      scale_fill_manual(values = c(
        "Expenses" = "brown1", "Revenue" = "forestgreen", "Net" = "black"
      )) +

      # set the colors manually

      facet_wrap(~Team) +

      # facet wrap by team
      
      scale_y_continuous(labels = number_format(scale = 1)) +
      
      # change axis so that it doesn't discplay scientific notation. I got advice 
      # during demo day that this would make my graphs more intuitive 

      labs(
        x = "Fiscal Year",
        y = "Amount in U.S. dollars",
        title = "Annual Team Expenses and Revenue",
        caption = "Data from U.S. soccer annual reports"
      ) +

      # relabel x and y, add title and caption

      theme_minimal() +

      # set theme to minimal

      theme(
        plot.title = element_text(hjust = 0.5)
      )

    # center title
```


# How much are female World Cup athletes paid around the world?

The question of equal (or unequal) investment in women's and men's soccer goes beyond the USWNT lawsuit. Please visit this project's [shiny app](https://bernadette-stadler.shinyapps.io/State-of-womens-soccer/) for more information on the state of women's soccer worldwide. 

# Is gender equality correlated with performance?

If paying male and female soccer players equally increases the performance of women's teams, that fact should incentivize national soccer governing bodies to do so. However, the real picture is more complicated, with gender equality (which I am using as a proxy for equal pay) only loosely correlated with better performance. GDP is much more strongly correlated with performance, suggesting it is not equality of investment between male and female teams that increases performance, but level of investment (as richer countries are able to invest more in their soccer teams).


```{r prep model data, include=FALSE}
model_data2 <- model_data %>%
      mutate(gendergap_18 = suppressWarnings(as.numeric((global_gender_gap_ranking_2018)))) %>%

      # mutate fifa data to change the name of global_gender_gap_ranking_2018 to something
      # easier to use and also to make it numeric. I used suppress warnings here to quiet
      # the NA introduced by coercion warning

      mutate(fifa_19 = suppressWarnings(as.numeric((fifa_wf_ranking_2019)))) %>%

      # make fifa_19, same logic as above

      mutate(fem_players = suppressWarnings(
        as.numeric((female_players_playing_organized_football))
      )) %>%

      # make fem_players, same logic as above

      select(x1, population, gendergap_18, fifa_19, fem_players) %>%

      # select variables of interest

      drop_na(gendergap_18)

    # drop NAs in gendergap_18. I didn't drop all NAs because it decimated the dataset,
    # but wanted to drop NAs from the variable that I will join by

    gdp_clean <- gdp_data %>% select(country_name, indicator_name, x2018)

    # select variables of interst in gdp data


    gender_index_clean <- gender_index %>%
      select(country_name, indicator, subindicator_type, x2018) %>%
      filter(indicator == "Overall Global Gender Gap Index" & subindicator_type == "Rank")

    # select columns and rows of interest from global gender gap data

    merge_1 <- gdp_clean %>%
      left_join(gender_index_clean, by = "country_name") %>%
      mutate(GDP = x2018.x) %>%
      mutate(gender_rank = x2018.y) %>%
      select(country_name, GDP, gender_rank)

    # join gdp and gender gap data, mutate to give year columns more intuitive names

    merge_2 <- model_data2 %>%
      left_join(merge_1, by = c("gendergap_18" = "gender_rank")) %>%

      # join merged data above with fifa data

      mutate(GDP = log10(GDP)) %>%
      mutate(Population = log10(population)) %>%
      mutate("Gender Gap Rank" = gendergap_18) %>%

      # take log of GDP and population so the graph won't look so crazy

      select(fifa_19, Population, "Gender Gap Rank", GDP) %>%

      # select variables of interest

      pivot_longer(cols = Population:GDP, names_to = "Type2", values_to = "value")

    # pivot longer so that I will be able to filter by type in the next step
```

```{r models, echo= FALSE, warning=FALSE}

GDP <- merge_2 %>%
  
      filter(Type2 == "GDP") %>%
      
      # filter by interactive variable

      ggplot(aes(x = value, y = fifa_19)) +

      # make graph

      geom_point() +
      geom_smooth(formula = y ~ x, method = "lm") +
      labs(
        y = "FIFA Rank", 
        x = "GDP (log scale)"
      )

    # add scatterplot and regression layers

Pop <- merge_2 %>%
  
      filter(Type2 == "Population") %>%
      ggplot(aes(x = value, y = fifa_19)) + 
      geom_point() +
      geom_smooth(formula = y ~ x, method = "lm") +
      labs(
        y = "FIFA Rank",
        x = "Population (log scale)"
      )

Gen_eq <- merge_2 %>%
  
      filter(Type2 == "Gender Gap Rank") %>%
      ggplot(aes(x = value, y = fifa_19)) + 
      geom_point() +
      geom_smooth(formula = y ~ x, method = "lm") +
      labs(
        y = "FIFA Rank", 
        x = "Gender Gap Rank"
      )

plot_grid(GDP, Pop, Gen_eq)
```
