---
title: "State of Women's Soccer 2019"
author: "Bernadette Stadler"
date: "12/10/2019"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readr)
library(ggplot2)
library(plotly)
library(readxl)
library(janitor)
library(reshape2)
library(scales)
library(gt)
library(sf)
library(fs)
library(cowplot)
library(tidyverse)
```

```{r read in data, include=FALSE}
  rev_exp <- read_excel("raw-data/Revenue and expense data.xlsx") %>%
    clean_names()

  # read in U.S. soccer revenue and expenses dataset

  wws <- read_csv("raw-data/tabula-GSSS 2017.csv", col_types = cols(
    league = col_character(),
    country = col_character(),
    Season = col_character(),
    Sport = col_character(),
    Teams = col_double(),
    Players = col_double()
  )) %>% clean_names()

  # read in womens salaries data

  mws <- suppressWarnings(read_csv("raw-data/tabula-GSSS 2017_2.csv", col_types = cols(
    RANK = col_character(),
    TEAM = col_character(),
    LEAGUE = col_character()
  ))) %>%
    drop_na() %>%
    filter(!LEAGUE == "LEAGUE") %>%
    clean_names()

  # read in mens salaries data for 2017.Use suppress warnings because I got a warning
  # saying that a missing column name was filled in with X4 that I couldn't get to
  # to go away. Filter to get rid of multiple table headers from subsequent pages of
  # the document I scraped it from.

  mws_18 <- suppressWarnings(read_csv("raw-data/tabula-GSSS 2018.csv", col_types = cols(
    COUNTRY = col_character(),
    CONTINENT = col_character(),
    X4 = col_character()
  ))) %>% clean_names()

  # read in mens salaries data for 2018.Use suppress warnings for same reason as above

  bonuses <- read_excel("raw-data/World cup bonuses.xlsx") %>% clean_names()

  # read in bonuses data (manually compiled from Guardian article)

  model_data <- read_csv("raw-data/womens-football-2019-clean.csv",
    col_types = cols(
      X1 = col_character(),
      Population = col_number(),
      `Global Gender Gap Ranking 2018` = col_character(),
      `FIFA WF Ranking 2019` = col_character(),
      `Female players playing organized football` = col_character(),
      `licesned coaches` = col_character(),
      `licensed referees` = col_character(),
      `Women's soccer strategy?` = col_character(),
      `Mixed-gender football?` = col_character()
    )
  ) %>% clean_names()
  
  # read in fifa data 

  gdp_data <- read_csv("raw-data/GDP_data.csv",
    skip = 4,
    col_types = cols(
      .default = col_double(),
      `Country Name` = col_character(),
      `Country Code` = col_character(),
      `Indicator Name` = col_character(),
      `Indicator Code` = col_character(),
      `2019` = col_logical(),
      X65 = col_logical()
    )
  ) %>% clean_names()
  
  # read in world bank data 

  gender_index <- read_csv("raw-data/data.csv",
    col_types = cols(
      `Country ISO3` = col_character(),
      `Country Name` = col_character(),
      `Indicator Id` = col_double(),
      Indicator = col_character(),
      `Subindicator Type` = col_character(),
      `2006` = col_double(),
      `2007` = col_double(),
      `2008` = col_double(),
      `2009` = col_double(),
      `2010` = col_double(),
      `2011` = col_double(),
      `2012` = col_double(),
      `2013` = col_double(),
      `2014` = col_double(),
      `2015` = col_double(),
      `2016` = col_double(),
      `2018` = col_double()
    )
  ) %>% clean_names()
  
  # read in gender index data 
  
  reg_season_salary <- read_excel("raw-data/reg season salaries.xlsx") %>% clean_names()
```

In March 2019, the 28 members of the United States Women's National Soccer Team (USWNT) filed a class-action lawsuit against their employer, U.S. Soccer. The lawsuit accuses U.S. soccer of violating the Equal Pay Act and Title VII of the Civil Rights Act of 1964 by consistently paying USWNT athletes less than their male counterparts (players on the U.S. Men's National Team or USMNT). The lawsuit, and the subsequent dominance of the USWNT in the 2019 Women's World Cup, have drawn significant attention to the issue of equal pay in professional soccer. However, the USWNT and USMNT have different pay structures that makes it difficult to compare their earnings. This project seeks to explain the USWNT and USMNT compensation structures and earning potentials in a way that allows for useful comparisons. It will also go beyond the USWNT to examine how professional female soccer players are compensated throughout the world.

# Regular Season Pay

The USWNT and USMNT have different payment structures, which makes it different to compare their salaries. USWNT players make a base salary of $100,000 and earn bonuses of on average $4,950 for games they win. The actual amount they earn may depend on the FIFA rating of their opponent. The USMNT does not have a base salary, but players make a bonus of $5,000 for losing a game and up to $17,625 for a win, depending on the opponent's ranking. On average, USMNT players would make around $13,166 per game that they win. The USMNT and USWNT also play differing number of games per year, further complicating efforts to directly compare their salaries. The graph below shows how much the USMNT and USWNT would make if each team played 20 games, depending on the number of games won.

```{r prep graph data, include = FALSE }

graph_data1 <- reg_season_salary %>% 
      pivot_longer(cols = women:men, names_to = "team", values_to = "salary") %>% 
      
      # pivot data so that women or men appear as values in column "team" so that I 
      # can facet wrap by team later 
      
      mutate(team = if_else(team == "women", "Women", "Men")) %>%
      
      # change "men" and "women" to "Men" and "Women" so that they will display 
      # nicely in the legend 
  
      filter(games_won %in% c(0,10,20)) %>%
    
      # filter so that I can show three different outcomes in my static graphic. In my app
      # this is interactive
      
      mutate(games_won = case_when(games_won == 0 ~ "No Games Won", 
                                   games_won == 10 ~ "Ten Games Won", 
                                   TRUE ~ "All Games Won"))
  
      # Changes numbers to something that will make sense as a label in my graph 

graph_data1$games_won <- factor(
  graph_data1$games_won, levels = c("No Games Won", "Ten Games Won", "All Games Won")
  )

      # I manually reorded the chart.
```

```{r reg season graph, echo = FALSE}

graph_data1 %>%
      ggplot() +
      geom_col(aes(x = team, y = salary, fill = team), position = "dodge") +
  
      # make a ggplot! Set position to dodge so the columns appear next to
      # eachother 
      
      facet_wrap(~games_won) +
      
      # facet wrap by games won to make three separate graphs 
      
      scale_fill_manual(values = c("Women" = "red", "Men" = "purple4")) +
      
      # manually color the bars, using same color scheme as rest of men/women 
      # graphs 
  
      scale_y_continuous(labels = number_format(scale = 0.001)) +
  
      # put numbers in regular notation (not scientific)
      
      coord_flip() +
      
      # flip coordinates because a two-column graph looks better going 
      # horizontally
      
      theme_minimal() +
      
      # theme of my app! 
      
      theme(axis.text.y = element_blank()) + 
      
      # remove y axis text because fill color already identifies each bar
      
      labs(title = "Money earned per games won", 
           subtitle = "Assuming each team plays 20 games", 
           y = "Money Earned (in thousands of dollars)", 
           x = "", 
           caption = "Data from USWNT lawsuit", 
           fill = "Team") +
  
      # add labels 
  
      theme(plot.title = element_text(hjust= 0.5)) +
      theme(plot.subtitle = element_text(hjust = 0.5))

      # center title and subtitle 


```


# World Cup Pay 

USMNT and USWNT players also make different amounts in bonuses for each stage of the world cup they reach, from being named to the world cup team to winning the world cup. The different bonuses made at each stage add up to quite a difference. This chart shows the cumulative amount that a USWNT or USMNT player would walk away with if their team exited the tournament at each respective stage.

```{r read in and clean data, include = FALSE}

bonuses <- read_excel("raw-data/World cup bonuses.xlsx") %>% clean_names()

# read in data, clean names

bonuses2 <- bonuses %>%
  select(stage, women, men, cumulative_women, cumulative_men) %>%

  # select columns of interest

  pivot_longer(
    cols = c("women", "men"), names_to = "team", values_to = c("Value")
  ) %>%

  # pivot longer so that men and women are in the column team so that I will
  # be able to facet wrap by team later to make two side-by-side graphc

  pivot_longer(cols = c("cumulative_women", "cumulative_men"), names_to = "team2", values_to = "cumulative") %>%

  # pivot longer again to get all of the cumulative data in one column.
  # unfortunatey this second pivot gave me some wierd rows that had both women
  # and men_cumulative or vice versa. So, see next step ...

  mutate(keep = case_when(
    team == "men" & team2 == "cumulative_women" ~ FALSE,
    team == "women" & team2 == "cumulative_men" ~ FALSE,
    TRUE ~ TRUE
  )) %>%

  # make a new column that identifies which rows mix teams with FALSE.

  filter(keep == "TRUE") %>%

  # filter for the rows that did not mixt teams, which I labeled as TRUE
  # in the last step

  select(stage, team, Value, cumulative) %>%

  # select variables of interest

  mutate(team = if_else(team == "women", "USWNT", "USMNT")) %>%

  # change "men" and "women" to "USWNT" and "USMNT" to make the graph look more
  # professional

  mutate(number = case_when(
    stage == "Start" ~ 1,
    stage == "Qualifying" ~ 2,
    stage == "Winning qualification games" ~ 3,
    stage == "Named to world cup team" ~ 4,
    stage == "Advance to knockout" ~ 5,
    stage == "Winning world cup" ~ 6,
    TRUE ~ 7
  ))

# give each stage a number. This allows me to make a line chart below (I was
# having some issues when I was using the discrete charcter values). 

```

```{r bonsuses graph, echo=FALSE}

ggplot(bonuses2, aes(x = number, y = cumulative)) +
  
  # make the ggplot and save it as world_cup_bonuses
  
  geom_point() +
  
  # add the geom_point layer
  
  geom_line() +
  
  # agg the geom_line layer 
  
  facet_wrap(~team) +
  
  # facet_wrap by team 
  
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7), labels = c("Start", "Qualifying", "Winning qualification games", "Named to world cup team", "Advance to knockout", "Winning world cup", "Victory tour")) +
  
  # manually add breaks at every number (previously it didn't display all of them),   # label each number with its associated stage 
  
  scale_y_continuous(labels = number_format(scale = 1)) +
  
  # change y axis labels so they disply as regular numbers, not scientific notation
  
  labs(x = "Stage",
       y = "Cumulative", 
       title = "Cumulative Bonuses per World Cup Stage", 
       caption = "Data from The Guardian") +
  
  # add labels to x and y axis, title, caption
  
  theme_minimal() +
  
  # theme minimal because that is the style of my app
  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

  # tilt tick labels for readability 

```

# Market realities? 

In response to the USWNT lawsuit, U.S. Soccer has claimed that it is justified in paying female athletes less because of the market reality that women's soccer generates less revenue. This graph shows USWNT and USMNT revenue, expenses, and net income for fiscal year 2014 through the projected values for 2019. While the USWNT had generated less income for U.S. soccer in some years, the reverse is true in others.

```{r}
rev <- rev_exp %>%

      # first, I'm just going to work with revenue data

      select(fiscal_year, womens_revenue, mens_revenue) %>%

      # select variables of interest

      pivot_longer(
        cols = ends_with("revenue"),
        names_to = "Team",
        values_to = "Revenue"
      ) %>%

      # pivot the data so that the USMNT/USWNT data ends up in one column
      # this will allow me to facet wrap my graph by men and women later

      mutate(fiscal_year = if_else(
        fiscal_year == "2019 (projected)", 2019, suppressWarnings(as.double((fiscal_year)))
      )) %>%

      # mutate to remove (projected) so that I can make fiscal_year a
      # double column. Use suppressWarnings to get rid of the NA introduced by coercion
      # warning (I couldn't figure out how to do it any other way)

      mutate(Team = if_else(Team == "womens_revenue", "USWNT", "USMNT"))

    # mutate to change womens_revenue and mens_revenue to USWNT and USMNT,
    # which is how I want it to show up in the graph

    exp <- rev_exp %>%

      # now work with expense data

      select(fiscal_year, womens_expenses, mens_expenses) %>%

      # select variables of interest

      pivot_longer(
        cols = ends_with("expenses"),
        names_to = "Team",
        values_to = "Expenses"
      ) %>%

      # pivot the data so that the USMNT/USWNT data ends up in one column
      # this will allow me to facet wrap my graph by men and women later

      mutate(fiscal_year = if_else(
        fiscal_year == "2019 (projected)", 2019, suppressWarnings(as.double(fiscal_year))
      )) %>%

      # make 2019 (projected) into a numeric value, as well as whole column. Use
      # suppress warning for same purposes as above.

      mutate(Team = if_else(Team == "womens_expenses", "USWNT", "USMNT"))

    # mutate mens_expenses and womens_expenses to USWNT and USMNT

    rev_exp_formatted <- exp %>%
      left_join(rev, by = c("fiscal_year", "Team")) %>%

      # rejoin the data

      mutate(Expenses = if_else(is.na(Expenses), 0, Expenses)) %>%

      # change NAs to 0 revenue so that I can collapse the rows. The NAs were
      # introduced because I have separate rows for revenue and expense

      mutate(Revenue = if_else(is.na(Revenue), 0, Revenue)) %>%

      # change NAs to 0 in revenue so that I can add and subtract data below

      mutate(Net = Revenue - Expenses) %>%

      # make a net column by subtracting expenses from revenue

      pivot_longer(
        cols = Expenses:Net,
        names_to = "Type",
        values_to = "Amount"
      )

    # pivot longer so that I will be able to filter by Revenue, Expenses and Net.
    # I'm not sure if this is by design, but this is one of the few ways I was able
    # to get reactive variables to work in shiny

    rev_exp_formatted$Type <- factor(rev_exp_formatted$Type, levels = c("Revenue", "Expenses", "Net"))

    # assign levels to Revenue, Expenses and Net so that they display with Net last
    # instead of in alphabetical order

    rev_exp_formatted %>%
      # make a ggplot

      ggplot(aes(x = fiscal_year, y = Amount, fill = Type)) +

      # fiscal_year goes on x axis, amount on y axis, color by type

      geom_col(position = "dodge") +

      # set columns to dodge so that they display side by side instead of stacked

      scale_fill_manual(values = c(
        "Expenses" = "brown1", "Revenue" = "forestgreen", "Net" = "black"
      )) +

      # set the colors manually

      facet_wrap(~Team) +

      # facet wrap by team
      
      scale_y_continuous(labels = number_format(scale = 1)) +
      
      # change axis so that it doesn't discplay scientific notation. I got advice 
      # during demo day that this would make my graphs more intuitive 

      labs(
        x = "Fiscal Year",
        y = "Amount in U.S. dollars",
        title = "Annual Team Expenses and Revenue",
        caption = "Data from U.S. soccer annual reports"
      ) +

      # relabel x and y, add title and caption

      theme_minimal() +

      # set theme to minimal

      theme(
        plot.title = element_text(hjust = 0.5)
      )

    # center title
```

# Equality in Professional Soccer Beyond the United States 

The question of equal (or unequal) investment in women's and men's soccer goes beyond the USWNT lawsuit. This section explores investment in men's and women's soccer at the professional level throughout the world.

# How much are female World Cup athletes paid around the world?
This data was taken from a New York Times article by Allison McCann, who conducted a survey of 108 female world cup athletes. Numerous countries are not represented; however, the map below gives a sense of the differences in salaries between female World Cup players. This map shows average values, but the minimum reported by players in McCann's survey was $0 (Jamaica) and the maximum was $350,000 (USA).

```{r read in map data}

salary_data <- read_excel("raw-data/Salaries worldwide.xlsx", col_names = c("salary", "country")) %>% clean_names()

# read in salary data, naming the columns appropriately 

untar("raw-data/ne_50m_admin_0_countries.zip")

# untar zip file containing world map shape file 

map3 <- st_read("ne_50m_admin_0_countries.shp") %>%
  mutate(NAME_EN = as.character(NAME_EN))

# save shape file as a r object. Change the column NAME_EN to character. 
# This is because I will be joining this data with the salary data later, 
# and the column I will be joinig it with is class character. Making this 
# change clears up some warnings, and this seemed like the best place 
# to do it. 

file_delete("ne_50m_admin_0_countries.README.html")

file_delete("ne_50m_admin_0_countries.VERSION.txt")

file_delete("ne_50m_admin_0_countries.cpg")

file_delete("ne_50m_admin_0_countries.dbf")

file_delete("ne_50m_admin_0_countries.prj")

file_delete("ne_50m_admin_0_countries.shx")

file_delete("ne_50m_admin_0_countries.shp")

# delete unnecessary files 

```

```{r manipulate map data, echo = FALSE}

salary_data %>% 
  mutate(salary = parse_number(salary)) %>%
  
  # convert salary to a number so I can manipulate it below. 
  # I will ultimatley be creating a column for average salary
  # for each country. 
  
  group_by(country) %>% 
  mutate(sum_salary_country = sum(salary)) %>% 
  ungroup() %>% 
  
  # create a column that sums the salaries for each country. 
  # To do this, I first grouped by country before using mutate.
  # Then I ungrouped the data because I have to group it 
  # differently for the next step
  
  group_by(country, sum_salary_country) %>% 
  count(country) %>% 
  mutate(avg_salary = sum_salary_country/n) %>%
  ungroup() %>% 
  
  # Group by country and sum_salary_country. Grouping by 
  # sum_salary_country is important because that keeps the
  # column in the data. Use count to find the number of times
  # each country appears in the data. Then make a new column, 
  # avg_salary that holds the average salary per country (found 
  # by dividing sum_salary_country by the number of times (n)
  # each country appeared in the data 

  select(-n, -sum_salary_country) %>% 
  
  # drop unnecessary columns 
  
  mutate(country = if_else(country == "United States", "United States of America", country))

  # update country column so that the label for the United States matches the shape 
  # file data 

map4 <- map3 %>% left_join(data_map, by = c("NAME_EN" = "country")) 

  # join salary data with shape file data by country. This will make it super easy 
  # to shade countries according to average salray in the next step. 

```

```{r make map} 
map <- ggplot() +
  geom_sf(data = map4, aes(fill = avg_salary)) + 
  
  # make a ggplot, using geom_sf to draw the world map and setting fill to 
  # average salary 
  
  scale_fill_gradient(na.value = "white", low = "tomato", high = "red") +
  
  # change the NA countries to white -- this just looks better. Set low to 
  # tomato and high to red because I've been using red throughout this project 
  # to represent women's teams. Tomato and Red make for a nice gradient. I did 
  # try using darker and lighter colors to increase contrast, but given the 
  # extreme difference between the highest and lowest values in this data, it 
  # made it look like there were two different colors on the map instead of a 
  # gradient. 
  
  labs(title = "Self-reported World Cup Salaries", 
       fill = "Average Salary", 
       caption = "Data from 'We Asked, They Answered: 108 Women's World Cup
       Players on Their Jobs, Money, and Sacrificing Everything'
       (New York Times Article by Allison McCann).") +
  
  # add title, retitle the legend, add a caption 
  
  theme_minimal() +
  
  # add minimal theme so it matches rest of app 
  
  theme(plot.caption = element_text(hjust = 0.5))

  # center the caption 
```

# Is gender equality correlated with performance?

If paying male and female soccer players equally increases the performance of women's teams, that fact should incentivize national soccer governing bodies to do so. However, the real picture is more complicated, with gender equality (which I am using as a proxy for equal pay) only loosely correlated with better performance. GDP is much more strongly correlated with performance, suggesting it is not equality of investment between male and female teams that increases performance, but level of investment (as richer countries are able to invest more in their soccer teams).


```{r}

```

